---
title: "prep_e18"
output: html_document
---

code to prep 10x e18 (mouse embryonic brain scATAC/scRNA multiome dataset from 10x genomics):
rna information from "e18.4.20210917.rds"
atac information from "e18_mouse_brain_fresh_5k_atac_fragments.tsv.gz"

@Nancy Zhang and @Sijia Huang for source of these two files or any preprocessing related to the original data.

```{r}
# calculate gene activity score by signac
# gene activity score is used by MaxFuse. Other methods directly use Fragments

library(Signac)

e18 = readRDS("/home/bkzhu/super_mario/atac_bench_nrz/10x_e18/data/e18.4.20210917.rds")
DefaultAssay(e18)='ATAC'
Annotation(e18)
frags=UpdatePath(Fragments(e18)[[1]], 
                 new.path = '/home/bkzhu/super_mario/atac_bench_nrz/10x_e18/data/e18_mouse_brain_fresh_5k_atac_fragments.tsv.gz')
Fragments(e18)=NULL
e18=SetAssayData(e18, slot = "fragments", new.data = frags)
gene.activities <- GeneActivity(e18)

# csnk2a1 duplication problem
temp = gene.activities[14703,] + gene.activities[14704,]
gene.activities = gene.activities[-c(14703, 14704),]
gene.activities = rbind(gene.activities,temp)
rownames(gene.activities)[21977] = 'Csnk2a1'
```


```{r}
## okay start saving out
e18_rna_sct = as.data.frame(t(e18@assays$SCT@data))
e18_rna_sct_names = colnames(e18_rna_sct)

library(Matrix)
## rna
e18_rna_sct = as(as.matrix(e18_rna_sct), "dgCMatrix")
writeMM(e18_rna_sct, "/home/bkzhu/super_mario/atac_bench_nrz/10x_e18/data/10x_e18_rna.txt")
write.csv(data.frame(names = e18_rna_sct_names), "/home/bkzhu/super_mario/atac_bench_nrz/10x_e18/data/10x_e18_rna_names.csv")

## atac_GAS
e18_gas = as.data.frame(t(gene.activities))
e18_gas_names = colnames(e18_gas)
e18_gas = as(as.matrix(e18_gas), "dgCMatrix")
writeMM(e18_gas, "/home/bkzhu/super_mario/atac_bench_nrz/10x_e18/data/10x_e18_GAS.txt")
write.csv(data.frame(names = e18_gas_names), "/home/bkzhu/super_mario/atac_bench_nrz/10x_e18/data/10x_e18_GAS_names.csv")

## atac_lsi
e18_lsi = e18@reductions$lsi@cell.embeddings[,c(2:50)]
write.csv(e18_lsi, "/home/bkzhu/super_mario/atac_bench_nrz/10x_e18/data/10x_e18_LSI49.csv")

## meta
meta_data = e18@meta.data
write.csv(meta_data, "/home/bkzhu/super_mario/atac_bench_nrz/10x_e18/data/10x_e18_meta.csv")
```


############## produce RNA and ATAC embedding for slt and ari calculation
## ATAC embedding can just direclty use LSI scores; RNA use PCA embedding; both 15 dimensions

```{r}
## rna
rna = readMM("/home/bkzhu/super_mario/atac_bench_nrz/10x_e18/data/10x_e18_rna.txt")
rna = as.matrix(rna)
rna_names = read.csv('/home/bkzhu/super_mario/atac_bench_nrz/10x_e18/data/rna_names.csv')
colnames(rna) = rna_names$names
rownames(rna) = paste0("cell",c(1:nrow(rna)))

library(Seurat)
temp_obj1 = CreateSeuratObject(counts=t(rna),assay="rna")
temp_obj1 = SetAssayData(object = temp_obj1, slot = "data", new.data = t(rna), assay="rna") # input data already sctnorm
temp_obj1 = ScaleData(temp_obj1)
temp_obj1 <- FindVariableFeatures(temp_obj1, selection.method = "vst", nfeatures = 2000)
temp_obj1 = RunPCA(temp_obj1, features = rownames(temp_obj1))
meta = read.csv("/home/bkzhu/super_mario/atac_bench_nrz/10x_e18/data/10x_e18_meta.csv")
pca = as.data.frame(temp_obj1@reductions$pca@cell.embeddings[,c(1:15)])
pca$label = meta$annotation
write.csv(pca, "/home/bkzhu/super_mario/atac_bench_nrz/10x_e18/data/orig_x.csv", row.names = F)

# lsi
lsi = read.csv("/home/bkzhu/super_mario/atac_bench_nrz/10x_e18/data/10x_e18_LSI49.csv")
lsi_orig = lsi[,c(1:15)]
lsi_orig$label = meta$annotation
write.csv(lsi_orig, "/home/bkzhu/super_mario/atac_bench_nrz/10x_e18/data/orig_y.csv", row.names = F)
```


