---
title: "prep_retina"
output: html_document
---

code to human retinal cells (scATAC/scRNA multiome dataset from https://www.sciencedirect.com/science/article/pii/S2666979X22001069 by Chang lab):
fragment files directly retrieved from GEO GSM5866073. (fragment files by cell type)
We first use Archr to calculate the gene activity scores (used for Maxfuse etc)


```{r}
library(ArchR)

AIIamacrine = "/home/bkzhu/super_mario/atac_bench_nrz/retina/data/atac/GSM5866073_AIIamacrine_frags.sort.tsv.gz"
Astrocyte = "/home/bkzhu/super_mario/atac_bench_nrz/retina/data/atac/GSM5866073_Astrocyte_frags.sort.tsv.gz"
Cone = "/home/bkzhu/super_mario/atac_bench_nrz/retina/data/atac/GSM5866073_Cone_frags.sort.tsv.gz"
GABAamacrine = "/home/bkzhu/super_mario/atac_bench_nrz/retina/data/atac/GSM5866073_GABAamacrine_frags.sort.tsv.gz"
Glyamacrine = "/home/bkzhu/super_mario/atac_bench_nrz/retina/data/atac/GSM5866073_Glyamacrine_frags.sort.tsv.gz"
Horizontal = "/home/bkzhu/super_mario/atac_bench_nrz/retina/data/atac/GSM5866073_Horizontal_frags.sort.tsv.gz"
Microglia = "/home/bkzhu/super_mario/atac_bench_nrz/retina/data/atac/GSM5866073_Microglia_frags.sort.tsv.gz"
Mullerglia = "/home/bkzhu/super_mario/atac_bench_nrz/retina/data/atac/GSM5866073_Mullerglia_frags.sort.tsv.gz"
OFFconebipolar = "/home/bkzhu/super_mario/atac_bench_nrz/retina/data/atac/GSM5866073_OFFconebipolar_frags.sort.tsv.gz"
ONconebipolar = "/home/bkzhu/super_mario/atac_bench_nrz/retina/data/atac/GSM5866073_ONconebipolar_frags.sort.tsv.gz"
Retinalganglioncell = "/home/bkzhu/super_mario/atac_bench_nrz/retina/data/atac/GSM5866073_Retinalganglioncell_frags.sort.tsv.gz"
Rodbipolar = "/home/bkzhu/super_mario/atac_bench_nrz/retina/data/atac/GSM5866073_Rodbipolar_frags.sort.tsv.gz"
Rod = "/home/bkzhu/super_mario/atac_bench_nrz/retina/data/atac/GSM5866073_Rod_frags.sort.tsv.gz"

paths = c(AIIamacrine,Astrocyte,Cone,GABAamacrine,Glyamacrine,Horizontal,Microglia,
          Mullerglia, OFFconebipolar, ONconebipolar, Retinalganglioncell, Rodbipolar, Rod)

names(paths) = c("AIIamacrine", "Astrocyte", "Cone", "GABAamacrine", "Glyamacrine", "Horizontal",
                 "Microglia", "Mullerglia", "OFFconebipolar", "ONconebipolar", "Retinalganglioncell",
                 "Rodbipolar", "Rod")

addArchRGenome("hg38")

ArrowFiles_retina <- ArchR::createArrowFiles(
  inputFiles = paths,
  sampleNames = names(paths),
  filterTSS = 4, #Dont set this too high because you can always increase later
  filterFrags = 1000, 
  addTileMat = TRUE,
  minFrags = 500,
  maxFrags = 1e+05,
  addGeneScoreMat = TRUE,
  nChunk = 2
)
ArrowFiles_retina
```


```{r}
# fast
proj_retina <- ArchRProject(
  ArrowFiles = ArrowFiles_retina, 
  outputDirectory = "/home/bkzhu/super_mario/atac_bench_nrz/retina/data/atac/",
  copyArrows = FALSE #This is recommened so that you maintain an unaltered copy for later usage.
)

#length(proj_retina$cellNames) # 50969 cells
```

```{r}
proj_retina <- addIterativeLSI(
    ArchRProj = proj_retina,
    useMatrix = "TileMatrix", 
    name = "IterativeLSI", 
    iterations = 2, 
    clusterParams = list( #See Seurat::FindClusters
        resolution = c(0.2), 
        sampleCells = 10000, 
        n.start = 10
    ), 
    varFeatures = 25000, 
    dimsToUse = 1:50
)

genescore_hub = getMatrixFromProject(
  ArchRProj = proj_retina,
  useMatrix = "GeneScoreMatrix",
  useSeqnames = NULL,
  verbose = TRUE,
  binarize = FALSE,
  threads = getArchRThreads(),
  logFile = createLogFile("getMatrixFromProject")
)

genescore_matrix_hub = t(genescore_hub@assays@data$GeneScoreMatrix)
genescore_matrix_hub = genescore_matrix_hub[proj_retina$cellNames,] # important
genescore_matrix_hub = as(genescore_matrix_hub, "dgCMatrix")
genescore_names_hub = getFeatures(proj_retina, useMatrix = "GeneScoreMatrix")

lsi = proj_retina@reducedDims$IterativeLSI@listData$matSVD

# make annotation
annotation = proj_retina$cellNames
annotation = gsub("#.*","",annotation)

# there are dup cells in the atac dataset, remove them
dupid = duplicated( gsub("^.*#","",proj_retina$cellNames))

##
gene_act_matx = genescore_matrix_hub[!dupid,]
lsi_49 = lsi[!dupid,-1]
annotation = annotation[!dupid]
table(annotation)
```


############# this part starts to prep rna #############
Files are standard 10x format, also retreived from the same GEO location


```{r}
data_dir <- '/home/bkzhu/super_mario/atac_bench_nrz/retina/data/rna/'
list.files(data_dir)
expression_matrix <- Read10X(data.dir = data_dir)
retina_rna_obj = CreateSeuratObject(counts = expression_matrix) # 51645 cells
```

```{r}
rna_barcode = colnames(retina_rna_obj)
rna_barcode = gsub("^.*_","",rna_barcode)
atac_barcode = proj_retina$cellNames
atac_barcode = gsub("^.*#","",atac_barcode)
atac_barcode = atac_barcode[!dupid]

matchid = match(atac_barcode, rna_barcode) #50008

# rearrange rna
retina_rna_obj_sort = retina_rna_obj[,matchid]
```


```{r}
# make multiome dataset
retina_multiome = CreateSeuratObject(counts = retina_rna_obj_sort@assays$RNA@counts, assay = "RNA")

colnames(gene_act_matx) = genescore_names_hub
rownames(gene_act_matx) = colnames(retina_multiome)
temp = CreateAssayObject(counts = t(gene_act_matx))
retina_multiome[["ATAC"]] = temp

# put in extra information
retina_multiome@meta.data$annotation = annotation
lsitemp = lsi[!dupid,]
rownames(lsitemp) = rownames(gene_act_matx)
colnames(lsitemp) <- paste0("LSI_", 1:50)
retina_multiome[["mds"]] <- CreateDimReducObject(embeddings = lsitemp, key = "LSI_", assay = DefaultAssay(retina_multiome))

# save out
saveRDS(retina_multiome, file = "/home/bkzhu/super_mario/atac_bench_nrz/retina/data/retina_seurat.rds")
```

Save out rna counts, gene activity score and LSI

```{r}
library(Matrix)

# rna related
writeMM(t(retina_multiome@assays$RNA@counts), "/home/bkzhu/super_mario/atac_bench_nrz/retina/data/rna.txt")
write.csv(data.frame(names = rownames(retina_multiome@assays$RNA@counts)),
          "/home/bkzhu/super_mario/atac_bench_nrz/retina/data/rna_names.csv", row.names = F)

# atac related
writeMM(t(retina_multiome@assays$ATAC@counts), "/home/bkzhu/super_mario/atac_bench_nrz/retina/data/atac.txt")
write.csv(data.frame(names = rownames(retina_multiome@assays$ATAC@counts)),
          "/home/bkzhu/super_mario/atac_bench_nrz/retina/data/atac_names.csv", row.names = F)
write.csv(lsi_49, "/home/bkzhu/super_mario/atac_bench_nrz/retina/data/lsi_49.csv") # forgot to remove first one so note in downstream stuff

# meta data
write.csv(retina_multiome@meta.data, "/home/bkzhu/super_mario/atac_bench_nrz/retina/data/meta.csv")
```


########## subsetting:
#### subset 20k cells, reason: slt and ari f1 score calculation can not handle more than 20k + 20k cells
## thus use 20k each for benchmarking in this case


```{r}
library(Matrix)

rna = readMM("/home/bkzhu/super_mario/atac_bench_nrz/retina/data/rna.txt")
atac = readMM("/home/bkzhu/super_mario/atac_bench_nrz/retina/data/atac.txt")
lsi = read.csv("/home/bkzhu/super_mario/atac_bench_nrz/retina/data/lsi_49.csv")
meta = read.csv("/home/bkzhu/super_mario/atac_bench_nrz/retina/data/meta.csv")

set.seed(42)
sub20k = sample(nrow(meta),20000)
#table(meta[sub20k,"annotation"])
writeMM(rna[sub20k,], "/home/bkzhu/super_mario/atac_bench_nrz/retina/data/rna_20k.txt")
writeMM(atac[sub20k,], "/home/bkzhu/super_mario/atac_bench_nrz/retina/data/atac_20k.txt")
write.csv(lsi[sub20k,],"/home/bkzhu/super_mario/atac_bench_nrz/retina/data/lsi_49_20k.csv", row.names = F)
write.csv(meta[sub20k,],"/home/bkzhu/super_mario/atac_bench_nrz/retina/data/meta_20k.csv", row.names = F)
```


############## produce RNA and ATAC embedding for slt and ari calculation
## ATAC embedding can just direclty use LSI scores; RNA use PCA embedding; both 15 dimensions

```{r}
## rna
rna = readMM("/home/bkzhu/super_mario/atac_bench_nrz/retina/data/rna_20k.txt")
rna = as.matrix(rna)
rna_names = read.csv('/home/bkzhu/super_mario/atac_bench_nrz/retina/data/rna_names.csv')
colnames(rna) = rna_names$names
rownames(rna) = paste0("cell",c(1:nrow(rna)))

library(Seurat)
temp_obj1 = CreateSeuratObject(counts=t(rna),assay="rna")
temp_obj1 = SetAssayData(object = temp_obj1, slot = "data", new.data = t(rna), assay="rna") # input data already sctnorm
temp_obj1 = ScaleData(temp_obj1)
temp_obj1 <- FindVariableFeatures(temp_obj1, selection.method = "vst", nfeatures = 2000)
temp_obj1 = RunPCA(temp_obj1, features = rownames(temp_obj1))
meta = read.csv("/home/bkzhu/super_mario/atac_bench_nrz/retina/data/meta_20k.csv")
pca = as.data.frame(temp_obj1@reductions$pca@cell.embeddings[,c(1:15)])
pca$label = meta$annotation
write.csv(pca, "/home/bkzhu/super_mario/atac_bench_nrz/retina/data/orig_x.csv", row.names = F)

# lsi
lsi = read.csv("/home/bkzhu/super_mario/atac_bench_nrz/retina/data/lsi_49_20k.csv")
lsi_orig = lsi[,c(2:16)]
lsi_orig$label = meta$annotation
write.csv(lsi_orig, "/home/bkzhu/super_mario/atac_bench_nrz/retina/data/orig_y.csv", row.names = F)
```

