---
title: "prep_cortical"
output: html_document
---

code to human cerebral cortical (scATAC/scRNA multiome dataset from https://pubmed.ncbi.nlm.nih.gov/34390642/ by greenleaf lab):
rna information from "Writeup14n_10x_greenleaf.RData"
atac information from "Writeup14n_10x_greenleaf.RData" # gene activity score pre calculated by signac already avaliable in the seurat object

@Nancy Zhang and @Sijia Huang for source of this file.

This processing code written by Zongming Ma

```{r}
load("Writeup14n_10x_greenleaf.RData")

dat1=t(greenleaf@assays$SCT@data)
dat2=t(greenleaf@assays$geneActivity@data)
names1= colnames(dat1)
names2= colnames(dat2)

temp= strsplit(names2, split="ATAC-")
temp=unlist(temp)
temp=temp[seq(2,length(temp),2)]
names2=temp

sum(names1 %in% names2)
sum(names2 %in% names1)

# make sure column names match
mm = match(names1, names2)
dat1=dat1[,!is.na(mm)]
dat2=dat2[,mm[!is.na(mm)]]

dat0=t(greenleaf@assays$RNA@data)
names0 = colnames(dat0)
names1 = colnames(dat1)
sum(names0 %in% names1)
sum(names1 %in% names0)
mm0 = match(names1, names0)
dat0=dat0[,mm0[!is.na(mm0)]]

dim(dat0)
# dim(dat1)
dim(dat2)
# make sure row names match
# rnacells=rownames(dat1)
rnacells=rownames(dat0)
ataccells=rownames(dat2)
plot(match(rnacells, ataccells))

# write out data matrices in sparse matrix form.
write.table(greenleaf@meta.data, file="greenleaf_cortical_meta.csv", sep=",", col.names=TRUE, row.names=FALSE)
# write out names of genes for RNA.
rna_names = as.data.frame(colnames(dat1))
colnames(rna_names)="names"
write.table(rna_names, file="greenleaf_cortical_rna_names.csv", sep=",", col.names=TRUE, row.names=TRUE)

# write out the sparse matrices.
writeMM(dat0, "greenleaf_cortical_RNAcount_data.mtx")
writeMM(dat1, "greenleaf_cortical_SCT_data.mtx")
writeMM(dat2, "greenleaf_cortical_GENEACTIVITY_data.mtx")

# calculate additional LSI scores
DefaultAssay(greenleaf) <- "ATAC"
greenleaf <- RunTFIDF(greenleaf)
greenleaf <- FindTopFeatures(greenleaf, min.cutoff = 'q0')
greenleaf <- RunSVD(greenleaf, n = 200)

dat3 = greenleaf@reductions$lsi@cell.embeddings
dim(dat3)
peakcells = rownames(dat3)
plot(match(rnacells, peakcells)) # make sure row names match.
write.table(dat3[,-1], file="greenleaf_cortical_peak_lsi.csv", sep=",", col.names=TRUE, row.names=TRUE)
```


############## produce RNA and ATAC embedding for slt and ari calculation
## ATAC embedding can just direclty use LSI scores; RNA use PCA embedding; both 15 dimensions

```{r}
## rna
rna = readMM("/home/bkzhu/super_mario/atac_bench_nrz/greanleaf_cortical/data/greenleaf_cortical_RNAcount_data.mtx")
rna = as.matrix(rna)
rna_names = read.csv('/home/bkzhu/super_mario/atac_bench_nrz/greanleaf_cortical/data/greenleaf_cortical_rna_names.csv')
colnames(rna) = rna_names$names
rownames(rna) = paste0("cell",c(1:nrow(rna)))

library(Seurat)
temp_obj1 = CreateSeuratObject(counts=t(rna),assay="rna")
temp_obj1 = SetAssayData(object = temp_obj1, slot = "data", new.data = t(rna), assay="rna") # input data already sctnorm
temp_obj1 = ScaleData(temp_obj1)
temp_obj1 <- FindVariableFeatures(temp_obj1, selection.method = "vst", nfeatures = 2000)
temp_obj1 = RunPCA(temp_obj1, features = rownames(temp_obj1))
meta = read.csv("/home/bkzhu/super_mario/atac_bench_nrz/greanleaf_cortical/data/greenleaf_cortical_meta.csv")
pca = as.data.frame(temp_obj1@reductions$pca@cell.embeddings[,c(1:15)])
pca$label = meta$annotation
write.csv(pca, "/home/bkzhu/super_mario/atac_bench_nrz/greanleaf_cortical/data/orig_x.csv", row.names = F)

# lsi
lsi = read.csv("/home/bkzhu/super_mario/atac_bench_nrz/greanleaf_cortical/data/greenleaf_cortical_peak_lsi.csv")
lsi_orig = lsi[,c(1:15)]
lsi_orig$label = meta$annotation
write.csv(lsi_orig, "/home/bkzhu/super_mario/atac_bench_nrz/greanleaf_cortical/data/orig_y.csv", row.names = F)
```






