---
title: "asap_dataprep"
output: html_document
---

asap prepping of the https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM4732109
(CD28 CD3 stim pbmc control group)

```{r}
adt = read.table("/home/bkzhu/super_mario/asap/files/GSM4732110_CD28_CD3_control_ASAP_ADT.tsv", header = TRUE)
# 300k cells
# there are more protein cells than atac cells. only take protein cells that are multiome
# done later on
```


#### directly process the fragment files (downloaded from the same geo depo location)
#### we process the fragment files by Archr


```{r}
# try to output a genescrore matrix from the archr
library(ArchR)
path1 = "/asap/files/GSM4732109_CD28_CD3_control_ASAP_fragments.tsv.gz"
addArchRGenome("hg38")
paths = c(path1)
names(paths) = c("pbmc")

ArrowFiles_asap_2 <- ArchR::createArrowFiles(
  inputFiles = paths,
  sampleNames = names(paths),
  filterTSS = 4, #Dont set this too high because you can always increase later
  filterFrags = 1000, 
  addTileMat = TRUE,
  minFrags = 500,
  maxFrags = 1e+05,
  addGeneScoreMat = TRUE,
  nChunk = 5,
  threads = 16,
  cleanTmp = TRUE
)

doubScores <- ArchR::addDoubletScores(
  input = ArrowFiles_asap_2,
  k = 10, #Refers to how many cells near a "pseudo-doublet" to count.
  knnMethod = "UMAP", #Refers to the embedding to use for nearest neighbor search.
  LSIMethod = 1
)

# fast
proj_asap2 <- ArchRProject(
  ArrowFiles = ArrowFiles_asap_2, 
  outputDirectory = "asap_archr",
  copyArrows = TRUE #This is recommened so that you maintain an unaltered copy for later usage.
)

cell_name_atac = gsub("pbmc#","",proj_asap2$cellNames)
sum(cell_name_atac %in% adt$barcode)


proj_asap2 <- addIterativeLSI(
    ArchRProj = proj_asap2,
    useMatrix = "TileMatrix", 
    name = "IterativeLSI", 
    iterations = 2, 
    clusterParams = list( #See Seurat::FindClusters
        resolution = c(0.2), 
        sampleCells = 10000, 
        n.start = 10
    ), 
    varFeatures = 25000, 
    dimsToUse = 1:50
)

proj_asap2 <- addUMAP(
    ArchRProj = proj_asap2, 
    reducedDims = "IterativeLSI", 
    name = "UMAP", 
    nNeighbors = 30, 
    minDist = 0.5, 
    metric = "cosine"
)

#### calculate the gene activity scores from the atac fragment files

genescore_asap2 = getMatrixFromProject(
  ArchRProj = proj_asap2,
  useMatrix = "GeneScoreMatrix",
  useSeqnames = NULL,
  verbose = TRUE,
  binarize = FALSE,
  threads = getArchRThreads(),
  logFile = createLogFile("getMatrixFromProject")
)
genescore_matrix_asap = t(genescore_asap2@assays@data$GeneScoreMatrix)
genescore_matrix_asap = as(genescore_matrix_asap, "dgCMatrix") 
genescore_names_asap = getFeatures(proj_asap2, useMatrix = "GeneScoreMatrix")
cell_names = colnames(genescore_asap2@assays@data$GeneScoreMatrix)
cell_names = gsub("pbmc#","",cell_names)

### save out the data

cell_names2 = proj_asap2$cellNames
cell_names2 = gsub("pbmc#","",cell_names2)
idx = match(cell_names, cell_names2)
lsi = proj_asap2@reducedDims$IterativeLSI@listData$matSVD
lsi_filt = lsi[idx,]
write.csv(lsi_filt, "/asap/lsi_pbmc_50.csv") # save out lsi embedding
writeMM(genescore_matrix_asap, "/asap/genescore_pbmc.txt") # save out gene scores
rnainfo = data.frame(names = genescore_names_asap)
write.csv(rnainfo, "/asap/genescore_names_pbmc.csv") # save out gene names
idx_pro = match(cell_names, adt$barcode)
write.csv(adt[idx_pro,], "/asap/adt_pbmc.csv") # save out the adt info from these cells
```


We will do clustering and manual annotation on these cells, using seurat v4 wnn clustering

```{r}
pro = read.csv("/wnn_local_related/asap/adt_pbmc.csv")
pro$X <- NULL
rownames(pro) = paste0("cell", c(1:nrow(pro)))
pro = t(pro)

library(Matrix)
atac = readMM("/wnn_local_related/asap/genescore_pbmc.txt")
atac_name = read.csv("/wnn_local_related/asap/genescore_names_pbmc.csv")
atac = as.data.frame(atac)
colnames(atac) = atac_name$names
rownames(atac) = paste0("cell", c(1:nrow(atac)))
atac = t(atac)

## create seurat v4 object
library(Seurat)
asapobj <- CreateSeuratObject(counts = pro, assay = "AB")
temp = CreateAssayObject(counts = atac, assay = "ATAC")
asapobj[["ATAC"]] = temp

DefaultAssay(asapobj) <- 'AB'
VariableFeatures(asapobj) <- rownames(asapobj[["AB"]])
asapobj <- NormalizeData(asapobj, normalization.method = 'CLR', margin = 2) %>% 
  ScaleData() %>% RunPCA(reduction.name = 'apca')

lsi = read.csv("/wnn_local_related/asap/lsi_pbmc_50.csv") # using LSI here as the atac embedding for WNN
lsi$X <- NULL
colnames(lsi) <- paste0("lsi_", 1:50)
rownames(lsi)=colnames(pro)
asapobj[["lsi"]] <- CreateDimReducObject(embeddings = as.matrix(lsi), key = "lsi_", assay = "ATAC")

```

# clustering on wnn embedding
```{r}
asapobj <- FindMultiModalNeighbors(asapobj, reduction.list = list("apca", "lsi"), dims.list = list(1:18, 2:40))
asapobj <- RunUMAP(asapobj, nn.name = "weighted.nn", reduction.name = "wnn.umap", reduction.key = "wnnUMAP_")
asapobj <- FindClusters(asapobj, graph.name = "wsnn", algorithm = 3, verbose = FALSE, resolution = 0.3)
p1 <- DimPlot(asapobj, reduction = 'wnn.umap', label = TRUE, repel = TRUE, label.size = 2.5) #+ NoLegend()
p1
```

# manual annotation
```{r}
new.cluster.ids <- c("Naive_CD4","Mem_CD4","Monocyte","NK","Naive_CD8","Mem_CD8","B","OtherT",
                     "dirt","Monocyte","Naive_CD4")
names(new.cluster.ids) <- levels(asapobj)
asapobj <- RenameIdents(asapobj, new.cluster.ids)
p3 <- DimPlot(asapobj, reduction = 'wnn.umap', label = TRUE) #+ NoLegend()
p3
```

# save out the meta data
```{r}
meta = read.csv("/wnn_local_related/asap/asap_pbmc_meta.csv")
meta$human_ann = as.character(Idents(asapobj))
write.csv(meta, "/wnn_local_related/asap/asap_pbmc_meta.csv")
```


And then we calculate the pca embedding for calc of slt and ari f1 score
for atac cells lsi embedding was used instead

```{r}
# script to make the orig embedding for benchmarking
# asap data
library(Matrix)
library(Seurat)
# pro
pro = read.csv("/home/bkzhu/super_mario/asap/data/adt_pbmc.csv")
rownames(pro) = as.character(c(1:nrow(pro)))
pro = pro[,c(-1,-2)] # remove
# lsi
lsi = read.csv("/home/bkzhu/super_mario/asap/data/lsi_pbmc_50.csv")
lsi = lsi[,c(-1,-2)]# remove the row index and first component
rownames(lsi) = as.character(c(1:nrow(lsi)))
# meta
meta = read.csv("/home/bkzhu/super_mario/asap/data/asap_pbmc_meta.csv")

## remove dirt
pro = pro[meta$human_ann != "dirt",]
lsi = lsi[meta$human_ann != "dirt",]

# use seurat as standard to produce reduction
x_obj=CreateSeuratObject(counts=t(pro),assay="x")
x_obj <- NormalizeData(x_obj)
x_obj <- ScaleData(x_obj, features = rownames(x_obj))
x_obj <- RunPCA(x_obj, features = rownames(x_obj)) #quick check elbow 15 is more then enough
pca_adt = as.data.frame(x_obj@reductions$pca@cell.embeddings[,c(1:15)])
pca_adt$label = meta$human_ann[meta$human_ann != "dirt"] #### could change if we want different labels
write.csv(pca_adt, "/home/bkzhu/super_mario/asap/data/orig_x.csv", row.names=FALSE)

# produce lsi reduction -> you dont really need to
lsi_peak = lsi[,c(1:15)]
lsi_peak$label = meta$human_ann[meta$human_ann != "dirt"] #### could change if we want different labels
write.csv(lsi_peak, "/home/bkzhu/super_mario/asap/data/orig_y.csv", row.names=FALSE)

```





