---
title: "codex_tonsil"
output: html_document
---

data prep script of codex tonsil data
source of the data:  https://onlinelibrary.wiley.com/doi/10.1002/eji.202048891
also described in: https://www.sciencedirect.com/science/article/abs/pii/S240547122100380X
tissue used (tonsil1 (as labeld in cell system paper) ID: tonsil9338 from Graham L. Barlow data transfer)

The processed single FOV mutli-channel tiffs (processed by internal codex servers) were directly transfered from nolan lab server.
The images then went through standard processing pipeline: whole cell segmentation (by mesmer with local implementation), signal extraction, signal compensation (with REDSEA), and that produces fcs files for each fov. These code were not supplied here since they are part of standard processing pipeline. 
Please contact Bokai Zhu if raw images from the original study, or processing code related to steps described above is needed.


For this study, we start with single cell codex data extracted in form of fcs files from REDSEA code: (tonsil 1)

```{r}
path = "/home/bkzhu/super_mario/tonsil_v2/Codex/FCS_output_DeepCell_extOnly/"
folders = list.files(path = path)
container = list()
i = 1
for (f in folders){
  fcs1=read.flowSet(paste0(path,f,'/WholeCell_REDSEA_RC=1/dataScaleSizeFCS.fcs'))
  expr1 = as.data.frame(fsApply(fcs1, exprs))
  expr1$PointNum = f
  container[[i]] = expr1
  i = i+1
}

library(dplyr)
exp_all = bind_rows(container, .id = "column_label") # 375,985
unique(exp_all$PointNum)
```

```{r}
t2u = c("cellLabelInImage","cellSize","HOECHST1","CD38","CD19","CD31","Vimentin","CD22","Ki67","CD8","CD90",
        "CD123","CD15","CD3","CD152","CD21","cytokeratin","CD2","CD66","collagen IV","CD81","HLA-DR",
        "CD57","CD4","CD7","CD278","podoplanin","CD45RA","CD34","CD54","CD9","IGM","CD117","CD56","CD279",
        "CD45","CD49f","CD5","CD16","CD63","CD11b","CD1c","CD40","CD274","CD27",
        "CD104","CD273","FAPalpha","Ecadherin","PointNum")
exp_all_s = exp_all[,t2u]

# with visual check roughly size should be around 30 - 800 pixels, fitler by cell size
exp_all_c = subset(exp_all_s, exp_all_s$cellSize>30 & exp_all_s$cellSize<800) # 371,085
```

```{r}
library(DelayedMatrixStats)
library(dplyr)
library(reshape)
library(ggplot2)


## here we do quantile norm to 0-1
expr1 = as.matrix(exp_all_c[,c(4:49)])
rng = colQuantiles(expr1, probs = c(0.05,0.95)) 
a = sweep(expr1, 2,rng[,1], "-")
b = rng[,2]-rng[,1]
c = sweep(a, 2, b, "/")
c[c>1]=1
c[c<0]=0
exp_all_c[,c(4:49)] = c # only the protein markers

```

Then we do a quick manual annotation of cell types:

```{r}
library(Seurat)
test = exp_all_c

temp_obj_ts <- CreateSeuratObject(counts = t(test[,c(4:49)]))
SetAssayData(object = temp_obj_ts, slot = "data", new.data = t(test[,c(4:49)])) # dont need to do addtional norm steps
temp_obj_ts <- SetAssayData(object = temp_obj_ts, slot = "scale.data", new.data = t(test[,c(4:49)]))
temp_obj_ts <- RunPCA(temp_obj_ts, features = colnames(test[,c(4:49)]))
ElbowPlot(temp_obj_ts)
temp_obj_ts <- RunUMAP(temp_obj_ts, dims = 1:15)
temp_obj_ts = FindNeighbors(temp_obj_ts , dims = 1:15)
temp_obj_ts = FindClusters(temp_obj_ts, resolution = 1)

temp_obj_ts_sub = subset(temp_obj_ts, downsample = 1000)
DimPlot(temp_obj_ts_sub, label = T)
```

```{r}
seurat_labels = Idents(temp_obj_ts)
# manual annotation
new.cluster.ids <- c("B-CD22-CD40","CD4 T","CD8 T","CD4 T","B-CD22-CD40","B-Ki67","B-Ki67",
                     "B-CD22-CD40","B-CD22-CD40","Fibro/Epi","Plasma","Other","Other","mix","Vessel",
                     "Fibro/Epi","Other","Bg","Fibro/Epi","Other","DC","Fibro/Epi","mix",
                     "Bg","mix","mix","Bg")
names(new.cluster.ids) <- levels(temp_obj_ts)
temp_obj_ts <- RenameIdents(temp_obj_ts, new.cluster.ids)
table(Idents(temp_obj_ts))

# remove the unwanted dirty cell, unannotated cells, and vessel cells that should not be included to mathc against rna cells
`%notin%` <- Negate(`%in%`)
exp_all_c_sub = subset(exp_all_c,exp_all_c$cluster.term %notin% c("mix","Vessel","Other","Bg")) #301,655

```

We dont need to run the whole tissue with 300k cells, instead we will just crop out roughly 7x8 FOV with ~ 150k cells
we want x 2 - 8 ; y 7 - 15:

```{r}
library(stringr)

fovs = c(as.character(unique(exp_all_c_sub$PointNum)))
list_fov = c()

for (i in c(2:8)){
  for (j in c(7:15)){
    x1 = str_sub(paste0("0",as.character(i)),-2,-1)
    y1 = str_sub(paste0("0",as.character(j)),-2,-1)
    target_xy = paste0('X',x1,'_','Y',y1)
    list_fov = c(list_fov, fovs[grep(target_xy,fovs)])
  }
}
list_fov
exp_all_c_sub_x28_y715 = subset(exp_all_c_sub, exp_all_c_sub$PointNum %in% list_fov)
# and save out
write.csv(exp_all_c_sub_x28_y715,"tonsil_v2/Codex/FCS_output_DeepCell_extOnly/formatch_clusters_x28_y715V2.csv")
```


