---
title: "prep_pbmc"
output: html_document
---

code to prep 10x pbmc (pbmc scATAC/scRNA multiome dataset from 10x genomics):
rna information from "pbmc_chromvar_annotated.rda"
atac information from "pbmc_chromvar_annotated.rda" # gene activity score pre calculated by signac already avaliable in the seurat object

@Nancy Zhang and @Sijia Huang for source of this file or any preprocessing related to the original data.


```{r}
load("/home/bkzhu/super_mario/atac_bench_nrz/10xpbmc/data/pbmc_chromvar_annotated.rda")

DefaultAssay(object = pbmc) <- "RNA"
# do sctransform as zmm requirement
pbmc <- PercentageFeatureSet(pbmc, pattern = "^MT-", col.name = "percent.mt")
pbmc[["SCT"]] <- NULL
pbmc <- SCTransform(pbmc, vars.to.regress = "percent.mt", verbose = FALSE) #reachediteration limit
```

```{r}
## okay start saving out
pbmc_rna_sct = as.data.frame(t(pbmc@assays$SCT@data))
pbmc_rna_sct_names = colnames(pbmc_rna_sct)

library(Matrix)
## rna
pbmc_rna_sct = as(as.matrix(pbmc_rna_sct), "dgCMatrix")
writeMM(pbmc_rna_sct, "/home/bkzhu/super_mario/atac_bench_nrz/10xpbmc/data/10x_pbmc_rna.txt")
write.csv(data.frame(names = pbmc_rna_sct_names), "/home/bkzhu/super_mario/atac_bench_nrz/10xpbmc/data/10x_pbmc_rna_names.csv")
## atac_GAS
pbmc_gas = as.data.frame(t(pbmc@assays$ACTIVITY@data))
pbmc_gas_names = colnames(pbmc_gas)
pbmc_gas = as(as.matrix(pbmc_gas), "dgCMatrix")
writeMM(pbmc_gas, "/home/bkzhu/super_mario/atac_bench_nrz/10xpbmc/data/10x_pbmc_GAS.txt")
write.csv(data.frame(names = pbmc_gas_names), "/home/bkzhu/super_mario/atac_bench_nrz/10xpbmc/data/10x_pbmc_GAS_names.csv")
## atac_lsi
pbmc_lsi = pbmc@reductions$lsi@cell.embeddings[,c(2:50)]
write.csv(pbmc_lsi, "/home/bkzhu/super_mario/atac_bench_nrz/10xpbmc/data/10x_pbmc_LSI49.csv")
## meta
meta_data = pbmc@meta.data
write.csv(meta_data, "/home/bkzhu/super_mario/atac_bench_nrz/10xpbmc/data/10x_pbmc_meta.csv")
```


############## produce RNA and ATAC embedding for slt and ari calculation
## ATAC embedding can just direclty use LSI scores; RNA use PCA embedding; both 15 dimensions

```{r}
## rna
rna = readMM("/home/bkzhu/super_mario/atac_bench_nrz/10xpbmc/data/10x_pbmc_rna.txt")
rna = as.matrix(rna)
rna_names = read.csv('/home/bkzhu/super_mario/atac_bench_nrz/10xpbmc/data/rna_names.csv')
colnames(rna) = rna_names$names
rownames(rna) = paste0("cell",c(1:nrow(rna)))

library(Seurat)
temp_obj1 = CreateSeuratObject(counts=t(rna),assay="rna")
temp_obj1 = SetAssayData(object = temp_obj1, slot = "data", new.data = t(rna), assay="rna") # input data already sctnorm
temp_obj1 = ScaleData(temp_obj1)
temp_obj1 <- FindVariableFeatures(temp_obj1, selection.method = "vst", nfeatures = 2000)
temp_obj1 = RunPCA(temp_obj1, features = rownames(temp_obj1))
meta = read.csv("/home/bkzhu/super_mario/atac_bench_nrz/10xpbmc/data/10x_pbmc_meta.csv")
pca = as.data.frame(temp_obj1@reductions$pca@cell.embeddings[,c(1:15)])
pca$label = meta$annotation
write.csv(pca, "/home/bkzhu/super_mario/atac_bench_nrz/10xpbmc/data/orig_x.csv", row.names = F)

# lsi
lsi = read.csv("/home/bkzhu/super_mario/atac_bench_nrz/10xpbmc/data/10x_pbmc_LSI49.csv")
lsi_orig = lsi[,c(1:15)]
lsi_orig$label = meta$annotation
write.csv(lsi_orig, "/home/bkzhu/super_mario/atac_bench_nrz/10xpbmc/data/orig_y.csv", row.names = F)
```




